---
title: "Data Visualization Assignment"
author: "Ching-Yi Cheng"
date: October 9 2026
format: 
    html:
      toc: true
      code-fold: true
      code-summary: "Show code"
      df-print: paged
      embed-resources: true
execute:
    echo: false
    warning: false
---

```{r}
library(tidyverse)
library(psych)
library(readxl)
library(gt)
library(janitor)
library(ggplot2)
library(readr)
library(dplyr)
library(scales)
raw_data <- read_csv("english_education.csv", show_col_types = FALSE)
```

# Introduction
This analysis uses Office for National Statistics data following state-school pupils who took their GCSEs in 2012-2013, with educational records extending from primary school to age 22. It focuses on small, medium, and large towns across eight English regions. Higher income deprivation indicates greater hardship linked to low income.

# Data Resource
Office for National Statistics. (2023). Educational attainment of young people in English towns data [Data set]. https://www.ons.gov.uk/peoplepopulationandcommunity/educationandchildcare/datasets/educationalattainmentofyoungpeopleinenglishtownsdata

## Version A: Scientific Article
```{r}
#Prepare variables
income_order <- c(
  "Higher deprivation towns",
  "Mid deprivation towns",
  "Lower deprivation towns"
)

scientific_data <- raw_data |>
  select(
    size_flag,
    rgn11nm,
    income_flag,
    education_score
  ) |>
  filter(
    size_flag %in% c(
      "Small Towns",
      "Medium Towns",
      "Large Towns"
    ),
    !is.na(rgn11nm),
    income_flag %in% income_order,
    !is.na(education_score)
  ) |>
  mutate(
    income_flag = factor(
      income_flag,
      levels = income_order
    )
  )
```

### 1.Box plot
Within all eight regions, median town education scores increase from higher to lower income deprivation categories. However, the score distributions overlap, indicating variation in educational attainment within each category.

```{r}
#| label: fig-education-boxplot
#| fig-width: 8
#| fig-height: 11
scientific_data |>
  ggplot(
    aes(
      x = income_flag,
      y = education_score,
      fill = income_flag
    )
  ) +
  geom_boxplot(
    width = 0.55,
    colour = "black",
    linewidth = 0.5,
    outlier.size = 1.5,
    outlier.alpha = 0.6,
    show.legend = FALSE
  ) +
  facet_wrap(
    vars(rgn11nm),
    ncol = 2
  ) +
  scale_x_discrete(
    labels = c(
      "Higher deprivation towns" = "Higher",
      "Mid deprivation towns" = "Mid",
      "Lower deprivation towns" = "Lower"
    ),
    drop = FALSE
  ) +
  scale_fill_manual(
    values = c(
      "Higher deprivation towns" = "#D95F5F",
      "Mid deprivation towns" = "#E6C84A",
      "Lower deprivation towns" = "#4C9BD2"
    )
  ) +
  labs(
    title = "Town educational attainment by income deprivation and region",
    x = "Income deprivation category",
    y = "Town education score",
  ) +
  theme_bw(
    base_size = 12
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(
      fill = "grey95",
      colour = "grey60"
    ),
    strip.text = element_text(
      face = "bold"
    ),
    plot.title.position = "plot",
    plot.caption = element_text(
      hjust = 0
    )
  )
```

### 2.Dumbbell plot
Lower-deprivation towns have higher mean education scores than higher-deprivation towns in every region, with each town weighted equally. The difference is largest in the North East and smallest in the South West, showing that the magnitude of the within-region disparity varies geographically.
```{r}
#Calculate mean education scores within each region and category
regional_education_summary <- scientific_data |>
  summarise(
    n_towns = n(),
    mean_education_score = mean(education_score),
    .by = c(
      rgn11nm,
      income_flag
    )
  )

#Calculate the difference between Lower and Higher categories
regional_education_gap <- regional_education_summary |>
  filter(
    income_flag %in% c(
      "Higher deprivation towns",
      "Lower deprivation towns"
    )
  ) |>
  select(
    rgn11nm,
    income_flag,
    mean_education_score
  ) |>
  pivot_wider(
    names_from = income_flag,
    values_from = mean_education_score
  ) |>
  mutate(
    education_gap =
      `Lower deprivation towns` -
      `Higher deprivation towns`
  ) |>
  filter(
    !is.na(education_gap)
  )

#Order regions by the difference
#The largest difference appears at the top
gap_region_order <- regional_education_gap |>
  arrange(education_gap) |>
  pull(rgn11nm)

#Prepare the two group means for plotting
regional_education_means <- regional_education_summary |>
  filter(
    income_flag %in% c(
      "Higher deprivation towns",
      "Lower deprivation towns"
    ),
    rgn11nm %in% gap_region_order
  ) |>
  mutate(
    rgn11nm = factor(
      rgn11nm,
      levels = gap_region_order
    ),
    income_flag = factor(
      income_flag,
      levels = c(
        "Higher deprivation towns",
        "Lower deprivation towns"
      )
    )
  )

#Draw the dumbbell plot
regional_education_means |>
  ggplot() +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    colour = "#BEBEBE",
    linewidth = 0.5
  ) +
  geom_segment(
    data = regional_education_gap,
    aes(
      x = `Higher deprivation towns`,
      xend = `Lower deprivation towns`,
      y = factor(
        rgn11nm,
        levels = gap_region_order
      ),
      yend = factor(
        rgn11nm,
        levels = gap_region_order
      )
    ),
    colour = "#BEBEBE",
    linewidth = 1
  ) +
  geom_point(
    aes(
      x = mean_education_score,
      y = rgn11nm,
      shape = income_flag,
      colour = income_flag,
      fill = income_flag
    ),
    size = 3.5,
    stroke = 1
  ) +
  scale_shape_manual(
    name = "Town income deprivation",
    values = c(
      "Higher deprivation towns" = 21,
      "Lower deprivation towns" = 19
    ),
    labels = c(
      "Higher deprivation towns" = "Higher deprivation",
      "Lower deprivation towns" = "Lower deprivation"
    )
  ) +
  scale_colour_manual(
    name = "Town income deprivation",
    values = c(
      "Higher deprivation towns" = "#FF5151",
      "Lower deprivation towns" = "#66B3FF"
    ),
    labels = c(
      "Higher deprivation towns" = "Higher deprivation",
      "Lower deprivation towns" = "Lower deprivation"
    )
  ) +
  scale_fill_manual(
    name = "Town income deprivation",
    values = c(
      "Higher deprivation towns" = "#FF5151",
      "Lower deprivation towns" = "#66B3FF"
    ),
    labels = c(
      "Higher deprivation towns" = "Higher deprivation",
      "Lower deprivation towns" = "Lower deprivation"
    )
  ) +
  scale_x_continuous(
    expand = expansion(
      mult = c(0.08, 0.08)
    )
  ) +
  labs(
    title = "Mean educational attainment by deprivation category",
    x = "Mean town education score",
    y = NULL
  ) +
  theme_bw(
    base_size = 12
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot",
    plot.caption = element_text(
      hjust = 0
    )
  )
```


## Version B: Accessible / Dynamic Outlet
### Prepare variables
```{r}
# Calculate the number of towns and mean score in each group
regional_education_summary <- scientific_data |>
  summarise(
    n_towns = n(),
    mean_education_score = mean(education_score),
    .by = c(
      rgn11nm,
      income_flag
    )
  )

# Calculate the difference between Lower and Higher categories
regional_education_gap <- regional_education_summary |>
  filter(
    income_flag %in% c(
      "Higher deprivation towns",
      "Lower deprivation towns"
    )
  ) |>
  select(
    rgn11nm,
    income_flag,
    mean_education_score
  ) |>
  pivot_wider(
    names_from = income_flag,
    values_from = mean_education_score
  ) |>
  mutate(
    education_gap =
      `Lower deprivation towns` -
      `Higher deprivation towns`
  ) |>
  filter(
    !is.na(education_gap)
  )
```


## Version B: Accessible / Dynamic Outlet
```{r}
#Select variables
data <- raw_data |>
  select(
    size_flag,
    rgn11nm,
    income_flag,
    education_score
  )
```

```{r}
#Prepare the new data frame "regional_income"
regional_income <- data |>
  filter(                     #Keep only relevant 
    size_flag %in% c(         #Include three town-size
      "Small Towns",
      "Medium Towns",
      "Large Towns"
    ),
    !is.na(rgn11nm),          #Remove towns with missing
    income_flag %in% c(       #Keep the three deprivation
      "Higher deprivation towns",
      "Mid deprivation towns",
      "Lower deprivation towns"
    )
  ) |>
  count(                      #Count towns by region and 
    rgn11nm,
    income_flag,
    name = "n_towns"
  ) |>
  mutate(                                 #Create region-level variables
    region_total = sum(n_towns),          #Calculate total towns per region
    proportion = n_towns / region_total,  #Calculate each category's share
    .by = rgn11nm                         #Perform calculations within each region
  )
```

```{r}
# Prepare the new data frame "regional_order"
region_order <- regional_income |>
  filter(
    income_flag == "Higher deprivation towns"
  ) |>
  arrange(proportion) |>
  pull(rgn11nm)

income_order <- c(
  "Higher deprivation towns",
  "Mid deprivation towns",
  "Lower deprivation towns"
)
```

### 1.Table
The North West has the most higher-deprivation towns: 93, compared with 38 in the South East. Regions contain different numbers of towns, so we next compare percentages to see where higher-deprivation towns are more common.
```{r}
income_table <- regional_income |>
  select(
    rgn11nm,
    income_flag,
    n_towns
  ) |>
  pivot_wider(
    names_from = income_flag,
    values_from = n_towns,
    values_fill = 0
  ) |>
  mutate(
    Total = `Higher deprivation towns` +
      `Mid deprivation towns` +
      `Lower deprivation towns`,
    rgn11nm = factor(
      rgn11nm,
      levels = rev(region_order)
    )
  ) |>
  arrange(rgn11nm)

income_table |>
  gt() |>
  cols_label(
    rgn11nm = "Region",
    `Higher deprivation towns` = "Higher",
    `Mid deprivation towns` = "Mid",
    `Lower deprivation towns` = "Lower",
    Total = "Total"
  ) |>
  tab_spanner(
    label = "Income deprivation category",
    columns = c(
      `Higher deprivation towns`,
      `Mid deprivation towns`,
      `Lower deprivation towns`
    )
  ) |>
  tab_header(
    title = "How many towns fall into each income deprivation category?",
    subtitle = "Number of towns in each English region"
  ) |>
  cols_align(
    align = "center",
    columns = -rgn11nm
  )
```

### 2.Stacked Bar Chart
The North East has the highest share of higher-deprivation towns, at about 77%, while Yorkshire and The Humber and the North West both have shares around 55%. These percentages compare towns rather than residents, raising the next question: do regions with larger shares also have lower average education scores?
```{r}
regional_income |>
  mutate(
    rgn11nm = factor(
      rgn11nm,
      levels = region_order
    ),
    income_flag = factor(
      income_flag,
      levels = income_order
    )
  ) |>
  ggplot(
    aes(
      x = rgn11nm,
      y = proportion,
      fill = income_flag
    )
  ) +
  geom_col(
    width = 0.90,
    position = "stack"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = label_percent(),
    breaks = seq(0, 1, by = 0.2),
    expand = expansion(
      mult = c(0, 0.02)
    )
  ) +
  scale_fill_manual(
    values = c(
      "Higher deprivation towns" = "#D95F5F",
      "Mid deprivation towns" = "#E6C84A",
      "Lower deprivation towns" = "#4C9BD2"
    )
  ) +
  labs(
    title = "Income deprivation is distributed differently across regions",
    subtitle = "Regions are ordered by the share of towns with higher deprivation",
    x = NULL,
    y = "Share of towns",
    fill = "Income deprivation category"
  ) +
  theme_minimal(
    base_size = 14
  ) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.title.position = "plot"
  )
```

### 3.Scatter plot
The North East has the highest higher-deprivation share and the lowest average town education score, but the North West and Yorkshire and The Humber have similar shares and very different scores. Regional deprivation shares therefore do not consistently track educational attainment; these comparisons describe patterns rather than establish cause and effect.
```{r}
regional_education <- data |>
  filter(
    size_flag %in% c(
      "Small Towns",
      "Medium Towns",
      "Large Towns"
    ),
    !is.na(rgn11nm),
    income_flag %in% income_order
  ) |>
  summarise(
    higher_proportion = mean(
      income_flag == "Higher deprivation towns"
    ),
    mean_education_score = mean(
      education_score,
      na.rm = TRUE
    ),
    .by = rgn11nm
  )

regional_education |>
  ggplot(
    aes(x = higher_proportion, y = mean_education_score)
  ) +
  geom_hline(
    yintercept = 0, linetype = "dashed", colour = "grey65"
  ) +
  geom_point(
    size = 3,
    colour = "#0000E3"
  ) +
  ggrepel::geom_text_repel(
    aes(label = rgn11nm),
    seed = 123,
    size = 4,
    max.overlaps = Inf
  ) +
  scale_x_continuous(
    labels = scales::label_percent(),
    breaks = seq(0, 1, by = 0.1)
  ) +
  labs(
    title = "Do regions with more income-deprived towns also have lower average educational attainment?",
    x = "Share of towns with higher income deprivation",
    y = "Mean town education score"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
  )

```

In summary, these findings highlight the importance of examining income deprivation and educational attainment both within and across regions. Higher-deprivation towns have lower average education scores within every region, but regional deprivation shares alone do not explain differences in overall educational attainment. These patterns could guide further government investigation into local conditions, such as school resources and access to educational opportunities, that may help explain differences between regions.
